Modeling site-specific amino-acid preferences deepens phylogenetic estimates of the divergence of viral proteins.

### 2018-03-25 manuscript/manuscript.tex
Jesse edits / comments

When studying a protein's evolutionary history, it is important to understand the timescale at which the evolutionary events have occurred. 
For example, understanding how long a particular viral lineage has been circulating informs hypotheses about population size, evolutionary rate, and possible previous hosts. Molecular phylogenetic is commonly used to make sure estimates of time since the divergence of two lineages. 
Under a molecular clock assumption, the branch lengths on a phylogenetic tree can be transformed into estimates of absolute time. 
However, it has been observed that these phylogenetic estimates often appear to estimate times since deep divergences that appear to be too recent. 
That is, when two nodes are separated by a very long branch, the phylogenetic estimate of this branch length often appears to be too short, which leads to an estimate of time that is too recent.

In the case of viruses, there is independent evidence that the estimates of these deep divergences are indeed too recent. 
The times since the divergence of lineages within lentiviruses, paramyxoviruses and filoviruses have been estimated by methods which rely on information outside of or in addition to the viral phylogenetic tree.  
These independent estimates are often orders of magnitude older than the estimates using the viral phylogenetic tree only. 
For example, there are multiple examples of filovirus integrations into the genomes of rodents. 
Using these viral integrations and the fossil dates of the rodent species tree, the divergence of two main filovirus groups has been estimated to be $>7$ millions years ago. 
This estimate stands in stark contrast to the phylogenetic estimate of filovirus divergence which estimates this event happened $\sim 10,000$ years ago. 
All together, this indicates that phylogenetic branch length estimation does truly have a systematic bias towards underestimation.

Branch length underestimation is due in part to inadequate modeling of purifying selection. 
During evolution, protein sites do not sample all 20 available amino-acids equally. 
Instead, proteins have site-specific amino-acid preferences which are necessary to maintain the structure and function of the protein. 
Failure to account for these constraints will lead to branch length underestimation because the model will assume that sequences which have been evolving for a very long period of time should be more diverged than is actually the case. 
Furthermore, the site-specific amino-acid preferences at one site are often dependent on the amino-acid identity at another site. 
As a result, the preferences ``shift" over time time due to the accumulation of substitutions in other parts of the protein.

Most phylogenetic substitution models are time-homogenous and site-independent: they do not take into account how sites may interact with each other or how preferences may shift over time. 
However, they can model purifying selection with varying degrees of complexity. 
One method is to allow the parameter controlling the rate of nonsynonymous to synonymous substitution, the dN/dS parameter, to vary among the protein sites. 
This variation accounts for the fact that constrained sites will have a lower rate of amino-acid substitution than mutationally constrained sites. 
However, the stationary state of this model, which describes the expected sequence composition after a very long evolutionary time, will be exactly the same as a model without any variation in the dN/dS parameter. 
Alternatively, models can have site-specific stationary states which explicitly acknowledge the protein's site-specific amino-acid preferences. 
Such models should estimate deep divergence times because model still expects relatively low sequence divergence even after a very long evolutionary time.

Here, we tested the effect of modeling site-specific amino-acid preferences though a model with a site-specific stationary state on an influenza hemagglutinin (HA) phylogenetic tree. 
Specifically, we used Experimentally Informed Codon Models (ExpCM's) defined by site-specific amino-acid preferences measured by a high-throughput functional assay called deep mutational scanning. 
We found that these ExpCM's estimated longer branches. 
We found that the extension in branch length due modeling purifying selection via a site-specific stationary state is largely independent of the the effect of modeling purifying selection through rate variation. 
However, our results make it clear that the site-specific amino-preferences of HA are constant across the entire tree. 
In particular, the branch lengthening is most pronounced on branches leading to the HA sequence that was used in the deep mutational scanning experiment that parameterized the ExpCM.
These results underscore the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how this selection shifts over time.

@@

However, the increased realism of mutation-selection models comes at the cost of an increased number of parameters. 
Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
For instance, a GY94+$\Gamma\omega$ model with the commonly used F3X4 stationary state has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and nine parameters describing the nucleotide composition of the stationary state.
However, mutation-selection models must additionally specify 19 parameters defining the amino-acid preferences for \emph{each} site (there are 20 amino acids whose preferences are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, or 9,500 parameters for a 500-residue protein.
It is challenging to obtain values for these amino-acid preference parameters in a maximum-likelihood framework without overfitting the data~\citep{rodrigue2013statistical}.
Here we will primarily use experimentally informed codon models (ExpCM's), which define the site-specific amino-acid preference parameters \textit{a priori} from deep mutational scanning experiments so that they do not need to be fit from phylogenetic data~\citep[see Methods and][]{bloom2014experimentally, hilton2017phydms, bloom2017identification}.
Because the amino-acid preference parameters in an ExpCM are obtained from experiments, the number of ExpCM free parameters that are fit from the phylogenetic data are similar to a non-site-specific substitution model.
An alternative strategy of obtaining the amino-acid preference parameters via Bayesian inference~\citep{lartillot2004bayesian, rodrigue2014site} is discussed in the last section of the Results.

@@

\ref{fig:decay} shows how different substitution models predict amino-acid sequence identity to decrease as a function of branch length using model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters, with an asymptotic sequence divergence of 0.059. 
While this predicted sequence divergence is higher than $\frac{1}{20} = 0.05$, due to redundant codon and nucleotide biases favoring certain amino acids, it intuitively feels to low. 
Like many proteins HA has a highly conserved structure and function that imposes constraints that cause most sites to sample only a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.

@@

However, mutation-selection models such as ExpCM's have site-specific stationary states.
They predict that different sites will have different asymptotic sequence identities (\ref{fig:decay})---a prediction that accords with the empirical observation that some sites are much more variable than others in alignments of highly diverged sequences.
For instance, \ref{fig:decay} shows that at sites such as 183 and 305 in the H1 HA, an ExpCM but not a GY94-style model predicts that the divergence will always be low. 
When sites with highly constrained amino-acid preferences such as these are common, an ExpCM can estimate a long branch length at modest sequence identities that a GY94 model might attribute to a shorter branch.

@@

\begin{figure}
\centerline{\includegraphics[width=0.9\textwidth]{figures/simulations}}
\caption{\label{fig:simulations}
\textbf{Branch lengths inferred on data simulated under a model with site-specific amino-acid preferences.} 
We simulated alignments along a HA phylogenetic tree (see Figure~\ref{fig:empirical_trees}) using an ExpCM parameterized by the actual site-specific amino-acid preferences for an H1 HA~\citep{doud2016accurate}.
We then inferred the branch lengths of this tree from the simulated alignments.
The inferred branch lengths for various models are plotted on the x-axis, and the actual branch lengths used in the simulations are on the y-axis.
We performed 10 simulations and inferences, and gray points show each inferred branch length from each simulation, and black points show the average of each branch length across simulations.
The grey dashed line at $y=x$ represents the behavior of an unbiased estimator. 
}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{figures/empirical_trees.pdf}}
\caption{\label{fig:empirical_trees}
\textbf{Effect of site-specific amino-acid preferences and $\Gamma\omega$ rate variation on HA branch length estimation.} 
The branch lengths of the HA tree are optimized using the indicated ExpCM or GY94 model. 
The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logo plots for 15 sites in HA; the full set of experimentally measured amino-acid preferences are in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}. 
The ExpCM's use amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
Circle denotes the H1 clade and triangle denotes the H3 clade.
The trees are midpoint rooted based on the tree inferred by \texttt{RAxML} using the GTRCAT model. 
}
\end{figure}

@@

Deep mutational scanning has been used to measure the amino-acid preferences of all sites in two different HAs.
One scan measured the site-specific amino-acid preferences of an H1 HA~\citep{doud2016accurate} and the other measured the preferences of an H3 HA~\citep{lee2018deep}.
The amino-acid preferences measured for these two HAs are shown in \ref{suppfig:prefs_doud} and 
\ref{suppfig:prefs_lee}.
The H1 and H3 HAs have only $\sim$42\% amino-acid identity, and so are separated by a large distance on the phylogenetic tree (see triangle and circle on \ref{fig:empirical_trees}).
As described in \citet{lee2018deep}, the amino-acid preferences clearly differ between the H1 and H3 HA at a substantial number of sites (these differences are apparent in a simple visual comparison of \ref{suppfig:prefs_doud} and 
\ref{suppfig:prefs_lee}, see site 33 as an example).
Therefore, we also created a third set of amino-acid preferences by averaging the measurements for the H1 and H3 HAs, under the conjecture that these averaged preferences might better describe the ``average'' constraint on sites across the full HA tree.
These three sets of HA amino-acid preferences define three different ExpCM's.

@@

\begin{figure}
\centerline{\includegraphics[width=0.99\textwidth]{figures/phylobayes.pdf}}
\caption{\label{fig:phylobayes}
\textbf{Models inferred from natural sequences have similar stationary states to models defined by experimental preferences and estimate similar branch lengths.}
(A) We fit an ExpCM(H1+H3 avg)+$\Gamma\omega$ and a pbMutSel to the full HA tree in \ref{fig:empirical_trees}. 
The pbMutSel amino-acid preferences are inferred from the natural HA sequences, while the ExpCM amino-acid preferences are experimentally measured and then rescaled by the stringency parameter in \ref{tab:empirical_data}. 
The pbMutSel preferences are more correlated with the re-scaled average of the H1 and H3 deep mutational scanning preferences than the individual re-scaled H1 and H3 deep mutational scanning preferences are to each other (Pearson's $r$: 0.74 versus 0.52). 
(B) The ExpCM(H1+H3 avg)+$\Gamma\omega$ and pbMutSel models estimated similar branch lengths when fit to the entire HA tree. 
Points denote branch lengths between all pairs of tips on the tree. 
Blue and orange denote branches that lead to the H1 and H3 deep mutational scanning reference sequences respectively. 
The \texttt{phydms} program implementing ExpCM's and the \texttt{PhyloBayes-MPI} program implementing pbMutSel models give branch lengths in different units, so to facilitate direct comparison between the models, we have normalized all branch lengths returned by each program by the length of the branches separating the earliest (A/South Carolina/1918) and latest (A/Solomon Islands/2006) seasonal human H1 sequences on the tree. 
}
\end{figure}

@@

\section*{Discussion}

Here, we tested the effect of models with site-specific stationary states on branch length estimation for the phylogenetic tree of highly diverged influenza HA sequences. 
We used site-specific stationary state models called ExpCM's, which are defined by amino-acid preferences measured by deep mutational scanning. 
We found that the ExpCM's estimated longer branches and did so independently from substitution rate variation via $\Gamma\omega$. 
We also found that the ExpCM's estimated branch lengths of a similar length to models which infer the stationary state from the natural sequences in a Bayesian framework.

@@

However, it is clear from our results that a major limitation of site-independent, time-homogenous models is their inability to capture epistatic interactions. 
The site-specific amino-acid preferences of a protein shift across time and models with a single stationary state across the entire tree cannot capture these dynamics. 
The ExpCM's defined by preferences measured in one of two highly diverged HA's had the largest effect on branches near the HA from the experiment. 
The local effect of the ExpCM's defined by a single preference set indicate that different parts of the tree have different stationary states. 
This point is underscored by the comparison of the ExpCM defined by the average preference set and the pbMutSel model, which infers the stationary state from the natural sequences. 
Even though the stationary state of the pbMutSel model is more representative of the global stationary state, the ExpCM with average preferences has a larger effect on the branches from the H1 and H3 sequences. 
Therefore, while it is clear that modeling site-specific amino-acid preferences is important, neither one of these models is able to capture how these preferences shift over time.

@@

\section*{Methods}

@@

The GY94 model is M0 variant of the Goldman-Yang model described by \citet{yang2000codon}. 
Specifically, the substitution rate $P_{xy}$ from codon $x$ to codon $y$ is 
\begin{equation}
\label{eq:GY94}
P_{xy} = 
\begin{cases}
  0 & \mbox{if $x$ and $y$ differ by more than one nucleotide,}\\
  \Phi_y & \mbox{if $\mathcal{A}\left(x\right) = \mathcal{A}\left(y\right)$ and $x$ is converted to $y$ by a single-nucleotide transversion,} \\
  \omega \Phi_{y} & \mbox{if $\mathcal{A}\left(x\right) \ne \mathcal{A}\left(y\right)$ and $x$ is converted to $y$ by a single-nucleotide transversion,} \\
  \kappa \Phi_y & \mbox{if $\mathcal{A}\left(x\right) = \mathcal{A}\left(y\right)$ and $x$ is converted to $y$ by a single-nucleotide transition,} \\
  \omega \kappa \Phi_{y} & \mbox{if $\mathcal{A}\left(x\right) \ne \mathcal{A}\left(y\right)$ and $x$ is converted to $y$ by a single-nucleotide transition,} \\
  -\sum\limits_{z \ne x} P_{xz} & \mbox{if $x = y$,}
  \end{cases}
\end{equation}
where $\mathcal{A}\left(x\right)$ is the amino-acid encoded by codon $x$, $\kappa$ is the transition-transversion rate, $\Phi_y$ is the equilibrium frequency of codon $y$, and $\omega$ is the relative rate of nonsynonymous and synonymous substitutions.
We define the codon frequency parameters, $\Phi_y$, using the ``corrected F3X4" method from~\citep{pond2010correcting}. 
This method calculates the $\Phi_y$ values from the empirical alignment frequencies but corrects for the exclusion of sequences with premature stop codons from the analysis.

@@

The ExpCM models used in this paper are the ones described in \citet{bloom2017identification}. 
Briefly, the rate of substitution $P_{r,xy}$ of site $r$ from codon $x$ to $y$ is 
\begin{equation}
\label{eq:ExpCM}
P_{r,xy} = Q_{xy} \times F_{r,xy}
\end{equation}
where $Q_{xy}$ is proportional to the rate of mutation from $x$ to $y$, and $F_{r,xy}$ is proportional to the probability that this mutation fixes.

@@

@@

@@

@@

>>>>

\jdbcomment{I have commented on the general structure of the Introduction, which I think is mostly OK. 
Add in references next, as this is crucial.
The Intro should frame the general problem and describe specifically what is already known.
Right now it does a bit too much of repeating your first two sections of Results.
In particular, there is some balance between how much goes in Intro and Results. 
Right now we have a lot of good background in Results, so don't get too repetitive in Intro.}

When studying a protein's evolutionary history, it is important to understand the timescale at which the evolutionary events have occurred. 
For example, understanding how long a particular viral lineage has been circulating informs hypotheses about population size, evolutionary rate, and possible previous hosts. Molecular phylogenetic is commonly used to make sure estimates of time since the divergence of two lineages. 
Under a molecular clock assumption, the branch lengths on a phylogenetic tree can be transformed into estimates of absolute time. 
However, it has been observed that these phylogenetic estimates often appear to estimate times since deep divergences that appear to be too recent. 
That is, when two nodes are separated by a very long branch, the phylogenetic estimate of this branch length often appears to be too short, which leads to an estimate of time that is too recent. 
\jdbcomment{I would make the first sentence more general than just proteins.
Second sentence: good idea to use viruses as examples, but try to make the sentence a bit more specific and less general---also, population size may not be relevant to estimates of deep divergence times.
For the last part, can we in fact cite general non-viral references about times being too recent? 
If you can't find several good ones, then just drop that and make first paragraph solely about divergence estimates and have next paragraph go into underestimation.}

In the case of viruses, there is independent evidence that the estimates of these deep divergences are indeed too recent. 
The times since the divergence of lineages within lentiviruses, paramyxoviruses and filoviruses have been estimated by methods which rely on information outside of or in addition to the viral phylogenetic tree.  
These independent estimates are often orders of magnitude older than the estimates using the viral phylogenetic tree only. 
For example, there are multiple examples of filovirus integrations into the genomes of rodents. 
Using these viral integrations and the fossil dates of the rodent species tree, the divergence of two main filovirus groups has been estimated to be $>7$ millions years ago. 
This estimate stands in stark contrast to the phylogenetic estimate of filovirus divergence which estimates this event happened $\sim 10,000$ years ago. 
All together, this indicates that phylogenetic branch length estimation does truly have a systematic bias towards underestimation. 
\jdbcomment{Good content, although obviously needs references.
Also, I think it is best to have at most one or two general sentences and then get right into specific examples.
So I think paragraph would be stronger if you just have one (or two) general sentences and then bring up at least two specific examples, and finally return to general with ``similar things have been seen in viruses X, Y, and Z (citing each).}

Branch length underestimation is due in part to inadequate modeling of purifying selection. 
During evolution, protein sites do not sample all 20 available amino-acids equally. 
Instead, proteins have site-specific amino-acid preferences which are necessary to maintain the structure and function of the protein. 
Failure to account for these constraints will lead to branch length underestimation because the model will assume that sequences which have been evolving for a very long period of time should be more diverged than is actually the case. 
Furthermore, the site-specific amino-acid preferences at one site are often dependent on the amino-acid identity at another site. 
As a result, the preferences ``shift" over time time due to the accumulation of substitutions in other parts of the protein.  
\jdbcomment{Start with: "The under-estimation of deep branches is known to occur in part due to...
Also, we need to make this paragraph about what has \emph{already} been shown. 
It is not clear how much of what you argue here is things you can clearly cite to previous papers versus your new results.
So overall content is good, but make this paragraph talk about prior literature with specific citations to specific points.}

Most phylogenetic substitution models are time-homogenous and site-independent: they do not take into account how sites may interact with each other or how preferences may shift over time. 
However, they can model purifying selection with varying degrees of complexity. 
One method is to allow the parameter controlling the rate of nonsynonymous to synonymous substitution, the dN/dS parameter, to vary among the protein sites. 
This variation accounts for the fact that constrained sites will have a lower rate of amino-acid substitution than mutationally constrained sites. 
However, the stationary state of this model, which describes the expected sequence composition after a very long evolutionary time, will be exactly the same as a model without any variation in the dN/dS parameter. 
Alternatively, models can have site-specific stationary states which explicitly acknowledge the protein's site-specific amino-acid preferences. 
Such models should estimate deep divergence times because model still expects relatively low sequence divergence even after a very long evolutionary time. 
\jdbcomment{Generally good (except for lack of citations), but again is last point the one you are making or something that you can clearly cite from literature? See comment after next paragraph.}

Here, we tested the effect of modeling site-specific amino-acid preferences though a model with a site-specific stationary state on an influenza hemagglutinin (HA) phylogenetic tree. 
Specifically, we used Experimentally Informed Codon Models (ExpCM's) defined by site-specific amino-acid preferences measured by a high-throughput functional assay called deep mutational scanning. 
We found that these ExpCM's estimated longer branches. 
We found that the extension in branch length due modeling purifying selection via a site-specific stationary state is largely independent of the the effect of modeling purifying selection through rate variation. 
However, our results make it clear that the site-specific amino-preferences of HA are constant across the entire tree. 
In particular, the branch lengthening is most pronounced on branches leading to the HA sequence that was used in the deep mutational scanning experiment that parameterized the ExpCM.
These results underscore the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how this selection shifts over time. 
\jdbcomment{Looking at this final paragraph, I would suggest structuring it and the two above it (last three) like this:
\begin{itemize}
\item Under-estimation can result from failure to model purifying selection.
\item Most models are homogenous / independent, and model by rate variation.
\item Some models use site-specific stationary states.
\item Here we use ExpCMs and find...
\end{itemize}
So overall, pretty good structure that you already have but don't put too much of your results into the Introduction.

The one thing that you need to add is a better description of deep mutational scanning. 
So I would say spend less time repeating your results and more time providing that background in the last paragraph (or an additional added paragraph).
}

@@

However, the increased realism of mutation-selection models comes at the cost of an increased number of parameters. 
Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
For instance, a GY94+$\Gamma\omega$ model with the commonly used F3X4 stationary state has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and nine parameters describing the nucleotide composition of the stationary state.
However, mutation-selection models must additionally specify 19 parameters defining the amino-acid preferences for \emph{each} site (there are 20 amino acids whose preferences are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, or 9,500 parameters for a 500-residue protein.
It is challenging to obtain values for these amino-acid preference parameters in a maximum-likelihood framework without overfitting the data~\citep{rodrigue2013statistical}.
Here we will primarily use experimentally informed codon models (ExpCM's), which define the site-specific amino-acid preference parameters \textit{a priori} from deep mutational scanning experiments so that they do not need to be fit from phylogenetic data~\citep[see Methods and][]{bloom2014experimentally, hilton2017phydms, bloom2017identification}.
Because the amino-acid preference parameters in an ExpCM are obtained from experiments, the number of ExpCM free parameters is similar to a non-site-specific substitution model.
An alternative strategy of obtaining the amino-acid preference parameters via Bayesian inference~\citep{lartillot2004bayesian, rodrigue2014site} is discussed in the last section of the Results.

@@

\ref{fig:decay} shows how different substitution models predict amino-acid sequence identity to decrease as a function of branch length using model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters, with an asymptotic sequence identity of 0.059. 
While this predicted sequence identity is higher than $\frac{1}{20} = 0.05$, due to redundant codon and nucleotide biases favoring certain amino acids, it is much lower than the pairwise identity of even the most diverged HAs in nature.
While it is of course possible that the identity of HAs in nature would become even lower given more time, it seems biochemically improbable that it would ever become as low as 0.059. 
The reason is that like many proteins HA has a highly conserved structure and function that imposes constraints that cause many sites to sample only a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.

@@

However, mutation-selection models such as ExpCM's have site-specific stationary states.
They predict that different sites will have different asymptotic sequence identities (\ref{fig:decay})---a prediction that accords with the empirical observation that some sites are much more variable than others in alignments of highly diverged sequences.
For instance, \ref{fig:decay} shows that at sites such as 183 and 305 in the H1 HA, an ExpCM but not a GY94-style model predicts that the identity will always be relatively high. 
When sites with highly constrained amino-acid preferences such as these are common, an ExpCM can estimate a long branch length at modest sequence identities that a GY94 model might attribute to a shorter branch.

@@

\begin{figure}
\centerline{\includegraphics[width=0.9\textwidth]{figures/simulations}}
\caption{\label{fig:simulations}
\textbf{Branch lengths inferred on data simulated under a model with site-specific amino-acid preferences.} 
We simulated alignments along a HA phylogenetic tree (see \ref{fig:empirical_trees}) using an ExpCM parameterized by the actual site-specific amino-acid preferences for an H1 HA~\citep{doud2016accurate}.
We then inferred the branch lengths of this tree from the simulated alignments.
The inferred branch lengths for various models are plotted on the x-axis, and the actual branch lengths used in the simulations are on the y-axis.
We performed 10 simulations and inferences, and gray points show each inferred branch length from each simulation, and black points show the average of each branch length across simulations.
The grey dashed line at $y=x$ represents the behavior of an unbiased estimator. 
}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{figures/empirical_trees.pdf}}
\caption{\label{fig:empirical_trees}
\textbf{Effect of site-specific amino-acid preferences and $\Gamma\omega$ rate variation on HA branch length estimation.} 
The branch lengths of the HA tree were optimized using the indicated ExpCM or GY94 model. 
The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logo plots for 15 sites in HA; the full set of experimentally measured amino-acid preferences are in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}. 
The ExpCM's use amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
Circle denotes the H1 clade and triangle denotes the H3 clade.
The trees are midpoint rooted based on the tree inferred by \texttt{RAxML} using the GTRCAT model. 
}
\end{figure}

@@

Deep mutational scanning has been used to measure the amino-acid preferences of all sites in two different HAs.
One scan measured the preferences of an H1 HA~\citep{doud2016accurate} and the other measured the preferences of an H3 HA~\citep{lee2018deep}.
The amino-acid preferences measured for these two HAs are shown in \ref{suppfig:prefs_doud} and 
\ref{suppfig:prefs_lee}.
The H1 and H3 HAs have only $\sim$42\% amino-acid identity, and so are separated by a large distance on the phylogenetic tree (see triangle and circle on \ref{fig:empirical_trees}).
As described in \citet{lee2018deep}, the amino-acid preferences clearly differ between the H1 and H3 HA at a substantial number of sites (these differences are apparent in a simple visual comparison of \ref{suppfig:prefs_doud} and 
\ref{suppfig:prefs_lee}; see site 33 as an example).
Therefore, we also created a third set of amino-acid preferences by averaging the measurements for the H1 and H3 HAs, under the conjecture that these averaged preferences might better describe the ``average'' constraint on sites across the full HA tree \jdbcomment{reference suppfig with the average logo plot here}.
These three sets of HA amino-acid preferences define three different ExpCM's.

@@

\begin{figure}
\centerline{\includegraphics[width=0.99\textwidth]{figures/phylobayes.pdf}}
\caption{\label{fig:phylobayes}
\textbf{Models inferred from natural sequences have similar stationary states to models defined by experimental preferences and estimate similar branch lengths.}
\jdbcomment{Move the general text that applies to the all the figure (e.g., the first two sentences below) to \emph{before} the (A) label.}
(A) We fit an ExpCM(H1+H3 avg)+$\Gamma\omega$ and a pbMutSel to the full HA tree in \ref{fig:empirical_trees}. 
The pbMutSel amino-acid preferences are inferred from the natural HA sequences, while the ExpCM amino-acid preferences are experimentally measured and then rescaled by the stringency parameter in \ref{tab:empirical_data}. 
The pbMutSel preferences are more correlated with the re-scaled average of the H1 and H3 deep mutational scanning preferences than the individual re-scaled H1 and H3 deep mutational scanning preferences are to each other (Pearson's $r$: 0.74 versus 0.52). 
(B) The ExpCM(H1+H3 avg)+$\Gamma\omega$ and pbMutSel models estimated similar branch lengths when fit to the entire HA tree. 
Points denote branch lengths between all pairs of tips on the tree. 
Blue and orange denote branches that lead to the H1 and H3 deep mutational scanning reference sequences respectively. 
The \texttt{phydms} program implementing ExpCM's and the \texttt{PhyloBayes-MPI} program implementing pbMutSel models give branch lengths in different units, so to facilitate direct comparison between the models, we have normalized all branch lengths returned by each program by the length of the branches separating the earliest (A/South Carolina/1918) and latest (A/Solomon Islands/2006) seasonal human H1 sequences on the tree. 
}
\end{figure}

@@

\section*{Discussion}
\jdbcomment{Add references to this section.
Also, I think we are mostly using ``preferences'' rather than ``stationary state'', so use the former except in cases where the latter clearly makes more sense (e.g., talking about math rather than model.
}
Here, we tested the effect of models with site-specific stationary states on branch length estimation for the phylogenetic tree of highly diverged influenza HA sequences. 
We used site-specific stationary state models called ExpCM's, which are defined by amino-acid preferences measured by deep mutational scanning. 
We found that the ExpCM's estimated longer branches and did so independently from substitution rate variation via $\Gamma\omega$. 
We also found that the ExpCM's estimated branch lengths of a similar length to models which infer the stationary state from the natural sequences in a Bayesian framework.

@@

However, it is clear from our results that a major limitation of site-independent, time-homogenous models is their inability to capture epistatic interactions. 
The site-specific amino-acid preferences of a protein shift across time and models with a single stationary state across the entire tree cannot capture these dynamics. 
The ExpCM's defined by preferences measured in one of two highly diverged HA's had the largest effect on branches near the HA from the experiment. 
The local effect of the ExpCM's defined by a single preference set indicate that different parts of the tree have different stationary states. 
This point is underscored by the comparison of the ExpCM defined by the average preference set and the pbMutSel model, which infers the stationary state from the natural sequences. 
Even though the stationary state of the pbMutSel model is more representative of the global stationary state \jdbcomment{maybe ``tree average'' rather than ``global''}, the ExpCM with average preferences has a larger effect on the branches from the H1 and H3 sequences. 
Therefore, while it is clear that modeling site-specific amino-acid preferences is important, neither one of these models is able to capture how these preferences shift over time.

\jdbcomment{In particular, you need to add literature about how preferences shift.
These can mostly be citations worked into above. 
In particular, there are two types of citations: papers about how preferences shift over time. 
This would include Hugh's paper and ones cited therein. 
You don't need to cite all general papers on epistasis, but just the ones that focus on this in terms of preferences.

Second, papers about the idea of modeling non-independence in phylogenetics.
I believe that Richard Goldstein and David Pollock have a recent \textit{Nature Ecology \& Evolution} paper on this---look it up and also pull in relevant citations that they make.
Also, there may be some paper in this area by Rodrigue and Lartillot.
There are also a few cited in the third-to-last paragraph of discussion here (\url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4626756/}).
}

@@

\clearpage
\section*{Methods}

@@

The GY94 model is M0 variant of the Goldman-Yang model described by \citet{yang2000codon}. 
Specifically, the substitution rate $P_{xy}$ from codon $x$ to codon $y$ is 
\begin{equation}
\label{eq:GY94}
P_{xy} = 
\begin{cases}
  0 & \mbox{if $x$ and $y$ differ by more than one nucleotide,}\\
  \Phi_y & \mbox{if $\mathcal{A}\left(x\right) = \mathcal{A}\left(y\right)$ and $x$ is converted to $y$ by a single-nucleotide transversion,} \\
  \omega \Phi_{y} & \mbox{if $\mathcal{A}\left(x\right) \ne \mathcal{A}\left(y\right)$ and $x$ is converted to $y$ by a single-nucleotide transversion,} \\
  \kappa \Phi_y & \mbox{if $\mathcal{A}\left(x\right) = \mathcal{A}\left(y\right)$ and $x$ is converted to $y$ by a single-nucleotide transition,} \\
  \omega \kappa \Phi_{y} & \mbox{if $\mathcal{A}\left(x\right) \ne \mathcal{A}\left(y\right)$ and $x$ is converted to $y$ by a single-nucleotide transition,} \\
  -\sum\limits_{z \ne x} P_{xz} & \mbox{if $x = y$,}
  \end{cases}
\end{equation}
where $\mathcal{A}\left(x\right)$ is the amino-acid encoded by codon $x$, $\kappa$ is the transition-transversion rate, $\Phi_y$ is the equilibrium frequency of codon $y$, and $\omega$ is the relative rate of nonsynonymous and synonymous substitutions.
We define the codon frequency parameters, $\Phi_y$, using the ``corrected F3X4" method from~\citet{pond2010correcting}. 
This method calculates the $\Phi_y$ values from the empirical alignment frequencies but corrects for the exclusion of sequences with premature stop codons from the analysis. 
\jdbcomment{Also add a sentence saying how under these F3X4 there are 9 parameters defining $\Phi_x$.}

@@

The ExpCM models used in this paper are the ones described in \citet{bloom2017identification}. 
Briefly, the rate of substitution $P_{r,xy}$ of site $r$ from codon $x$ to $y$ is 
\begin{equation}
\label{eq:ExpCM}
P_{r,xy} = Q_{xy} \times F_{r,xy}
\end{equation}
where $Q_{xy}$ is proportional to the rate of mutation from $x$ to $y$, $F_{r,xy}$ is proportional to the probability that this mutation fixes, and the diagonal elements $P_{xx}$ are set by $P_{xx} = -\sum_{z \ne x} P_{xz}$.

@@

\jdbcomment{Further increase the number of rows in the logo plots to make them larger. Since we aren't under length constraints for supplement, might as well make them readable. Maybe 8 or so rows instead?}

@@

\jdbcomment{This seems fine, except you should take the \emph{minimum} rather than the \emph{mean} divergence between these samples.
Did you actually take minimum and just have typo in your comment above?
}

@@

\jdbcomment{
It is not true that all elements of $\mathbf{D}$ are $< 0$.
In fact, the principle eigenvector of a irreducible and acyclcic stochastic matrix is 1, and since $\mathbf{D}$ is a stochastic matrix minus the identity matrix, that means that it has one eigenvalue that is 0 and the rest are $< 0$.
I'm not sure how all the math works out, but I know that it should end up being the case that $\left[\lim_{t \rightarrow \infty} e^{t \mathbf{P_r}}\right]_{xy} = p_{r,x}$ (see \url{https://en.wikipedia.org/wiki/Stochastic_matrix}).

So the correct formula for the asymptotic sequence divergence at site $r$ must be
\begin{equation}
\sum_a \sum_{x \in a} p_{r,x} \sum_{y \in a} p_{r, y}.
\end{equation}
You can see that this is true without even doing all the complicated matrix exponential limits: at stationary state, the probability of being $x$ at site $r$ is $p_{r,x}$.
}

@@

\jdbcomment{OK, sounds fine.}
